Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2019-12-17 write-up/fda-recurrent.tex

\begin{assumption}[Event process assumptions]
\label{assumption:events}\normalfont
We assume the following holds:
\begin{enumerate}[label=(E.\arabic*)]
\item\label{E1} The subsampling rate is both lower and upper bounded for all at-risk times; that is,~$0 < L < \pi_i (t) < U < \infty$ for all~$i=1,2,\ldots$ and~$t \in [0, \tau]$ such that $R_i (t) = 1$.
\item\label{E2} There exists a nonnegative definite matrix~$\Xi (\theta)$ such that
  \[
    n^{-1} \Xi_n (\theta) = n^{-1} \sum_{i=1}^n \int_0^\tau w_i (t;
    \theta) (1 - w_i (t; \theta)) \left[ \frac{h_i^{(1)}(t;
        \theta)}{h_i (t; \theta)} \right] \times  \left [
      \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top R_i (t)
    dt \overset{P}{\to} \Xi (\theta).
  \]
\item\label{E3} There exists~$M$ such that $|W_{i,j,t}| < M$ for
  all~$(i,j,t)$.
\item\label{E4} For all~$j,k$
\[
n^{-1} \sum_{i=1}^n \int_0^\tau \left | \frac{d^2}{d\theta_j
    d\theta_k} h_i (t;\theta_0) \right|^2 R_i (t) dt \overset{P}{\to}
C < \infty
\]
as $n \to \infty$.
\end{enumerate}
\end{assumption}

We also require several assumptions due to the truncation of the Karhunen-Lo{\`e}ve decomposition that represents~$X(s,t)$.
\begin{assumption}[Functional assumptions~\citep{Park2018}] \normalfont
\label{assumption:truncation}
The following assumptions are standard in prior work on longitudinal functional data analysis~\citep{Park2018, Yao2005, ChenMuller2012}:
\begin{enumerate}[label=(A.\arabic*)]
\item\label{A1} $X = \{ X(s, t) : (s,t) \in \mathcal{S} \times
  \mathcal{T} \}$ is a square integrable element of the $L^2 (
  \mathcal{S} \times \mathcal{T})$
\item\label{A2} The subsampling and conditional intensity rate functions~$f_y(T)$ are continuous and~$\sup |f_y(T)| < \infty$.
\item\label{A3} $\E[X(s,t) X(s^\prime, t) X(s,t^\prime)
  X(s^\prime, t^\prime) ] < \infty$ for
  each $s,s^\prime \in [0,\Delta]$ and~$0 < t, t^\prime < \tau$.
\item\label{A4} $\E[\|X(\cdot, t)\|^4] < \infty$ for each~$0< t < \tau$.
\end{enumerate}
\end{assumption}
Finally, for simplicity, we assume that there exists~$b^\star$ such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the true function~$\beta(t)$ sits in the span of the spline basis expansion. With the necessary assumptions stated, we can now state Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.
\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:events},
Assumption~\ref{assumption:truncation}, and~$\Delta$ known, for
large~$n$ the estimator~$\hat \theta_n$ is consistent; moreover,
\[
\sqrt{n} (\hat \theta - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
An estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i}
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process, standard logistic regression software can be used to fit the recurrent event model by specifying an offset equal to~$\log \pi_i (t)$. Not only this, Lemma~\ref{prop:optimal} shows weights~\eqref{eq:waage_weights} are optimal within a particular class of weighted estimating equations.

@@

Under assumption~\ref{assumption:truncation}, we next consider the statistical efficiency of our proposed estimator when compared to complete-data maximum likelihood estimation. While subsampling introduces additional variation, it may significantly reduce the overall computational burden. It is this trade-off that we next make precise. To start, we consider the subsampling rate~$\pi(t) = c \times h(t; \theta)$ for $c>0$, i.e., the subsampling rate proportional to the intensity function with time-independent constant~$c > 0$. Under this subsampling rate, the weight function~\eqref{eq:waage_weights} is equal to $c / (c+1)$. Under Lemma~\ref{lemma:simpleasym},
\[
\Xi (\theta) = \frac{c}{c+1} \int_0^\tau \frac{ h^{(1)} (t; \theta)
  h^{(1)} (t; \theta)^\top}{h (t; \theta)} dt = \frac{c}{c+1} \Sigma (\theta)
\]
where~$\Sigma(\theta)$ is the Fisher information of the complete-data maximum likelihood estimator.
Therefore the relative efficiency is~$c/(1+c)$. For an upper bound~$H = \max_{t \in (0,\tau)} h(t;\theta)$, if we set~$\pi (t) = c \times H$, then the relative efficiency can be lower bounded by~$c / (c+1)$.

@@

@@

>>>>

\begin{assumption}[Event process assumptions]
\label{assumption:events}\normalfont
We assume the following holds:
\begin{enumerate}[label=(E.\arabic*)]
\item\label{E1} The subsampling rate is both lower and upper bounded for all at-risk times; that is,~$0 < L < \pi_i (t) < U < \infty$ for all~$i=1,2,\ldots$ and~$t \in [0, \tau]$ such that $R_i (t) = 1$.
\item\label{E1} The subsampling rate is both lower and upper bounded for all ``at risk times''; that is,~$0 < L < \pi_i (t) < U < \infty$ for all~$i=1,2,\ldots$ and~$t \in [0,\tau]$.
\item\label{E2} There exists a nonnegative definite matrix~$\Xi (\theta)$ such that
  \[
    n^{-1} \Xi_n (\theta) = n^{-1} \sum_{i=1}^n \int_0^\tau w_i (t;
    \theta) (1 - w_i (t; \theta)) \left[ \frac{h_i^{(1)}(t;
        \theta)}{h_i (t; \theta)} \right] \times  \left [
      \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top R_i (t)
    dt \overset{P}{\to} \Xi (\theta).
  \]
\item\label{E3} There exists~$M$ such that $|W_{i,j,t}| < M$ for all~$(i,j,t)$.
\item\label{E4} For all~$j,k$
\[
n^{-1} \sum_{i=1}^n \int_0^\tau \left | \frac{d^2}{d\theta_j
    d\theta_k} h_i (t;\theta_0) \right|^2 R_i (t) dt \overset{P}{\to}
C < \infty
\]
as $n \to \infty$.
\end{enumerate}
\end{assumption}

We also require several assumptions due to the truncation of the Karhunen-Lo{\`e}ve decomposition that represents~$X(s,t)$.
\begin{assumption}[Functional assumptions~\citep{Park2018}] \normalfont
\label{assumption:truncation}
The following assumptions are standard in prior work on longitudinal functional data analysis~\citep{Park2018, Yao2005, ChenMuller2012}:
\begin{enumerate}[label=(A.\arabic*)]
\item\label{A1} $X = \{ X(s, t) : (s,t) \in \mathcal{S} \times
  \mathcal{T} \}$ is a square integrable element of the $L^2 (
  \mathcal{S} \times \mathcal{T})$
\item\label{A2} The subsampling and conditional intensity rate functions~$f_y(T)$ are continuous and~$\sup |f_y(T)| < \infty$.
\item\label{A3} $\E[X(s,t) X(s^\prime, t) X(s,t^\prime)
  X(s^\prime, t^\prime) ] < \infty$ for
  each $s,s^\prime \in [0,\Delta]$ and~$0 < t, t^\prime < \tau$.
\item\label{A4} $\E[\|X(\cdot, t)\|^4] < \infty$ for each~$0< t < \tau$.
\end{enumerate}
\end{assumption}
Finally, for simplicity, we assume that there exists~$b^\star$ such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the true function~$\beta(t)$ sits in the span of the spline basis expansion. With the necessary assumptions stated, we can now state Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.

Assumption~\ref{A3} is guaranteed by~\ref{E1} and~\ref{E3}, but is stated here to emphasize its importance for the marginal covariance procedure of~\cite{Park2018}. Finally, for simplicity, we assume that there exists~$b^\star$ such that~$\beta(t) = \mathbold{\phi} (t) b^\star$; that is, the true function~$\beta(t)$ sits in the span of the spline basis expansion. With the necessary assumptions stated, we can now state Lemma~\ref{lemma:simpleasym}, which provides asymptotic theory.

\begin{lemma} \normalfont
\label{lemma:simpleasym}
Under Assumption~\ref{assumption:events}, Assumption~\ref{assumption:truncation}, and~$\Delta$ known, for large~$n$ the estimator~$\hat \theta_n$ is consistent; moreover,
\[
\sqrt{n} (\hat \theta - \theta) \to N(0, \Xi (\theta)^{-1})
\]
where
\[
  \Xi (\theta) = \int_{0}^{\tau} w(s; \theta) \left[ \frac{h^{(1)}(s;
      \theta) \times  h^{(1)} (s;\theta)^{\top}}{h(s; \theta)} \right]
  ds.
\]
and~$\tau$ is the random censoring time of the event process.
\end{lemma}
An estimator for~$\Xi(\theta)$ is
\begin{equation}
\label{eq:fisher}
  \hat \Xi (\theta) = n^{-1} \sum_{i=1}^n \sum_{t \in \bfT_i \cup D_i}
  w_i(t;\theta) (1-w_i(t;\theta)) \left[ \frac{h_i^{(1)}(t;
      \theta)}{h_i (t; \theta)} \right] \times  \left [
    \frac{h_i^{(1)} (t;\theta)}{h_i(t; \theta)} \right]^\top
\end{equation}
For the log-linear intensity model, the sampling-unbiased estimator for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the previously described logistic regression model.
This implies that subsampling from an inhomogeneous Poisson process, standard logistic regression software can be used to fit the recurrent event model by specifying an offset equal to~$\log \pi_i (t)$. Not only this, Lemma~\ref{prop:optimal} shows weights~\eqref{eq:waage_weights} are optimal within a particular class of weighted estimating equations.

@@

Under assumption~\ref{assumption:truncation}, we next consider the statistical efficiency of our proposed estimator when compared to complete-data maximum likelihood estimation. While subsampling introduces additional variation, it may significantly reduce the overall computational burden. It is this trade-off that we next make precise. To start, we consider the subsampling rate~$\pi(t) = c \times h(t; \theta)$ for $c>0$, i.e., the subsampling rate proportional to the intensity function with time-independent constant~$c > 0$. Under this subsampling rate, the weight function~\eqref{eq:waage_weights} is equal to $c / (c+1)$. Under Lemma~\ref{lemma:simpleasym},

For the log-linear intensity model, the sampling-unbiased estimator for~$\hat \Xi(\theta)$ is equivalent to the Fisher information for the previously described logistic regression model. This implies that subsampling from an inhomogeneous Poisson process, standard logistic regression software can be used to fit the recurrent event model by specifying an offset equal to~$\log \pi_i (t)$. Not only this, Lemma~\ref{prop:optimal} shows weights~\eqref{eq:waage_weights} are optimal within a particular class of weighted estimating equations.

\begin{prop} \normalfont
\label{prop:optimal}
If the event process is an inhomogeneous Poisson point process with intensity~$h(t; \theta)$ and subsampling occurs via an independent, inhomogeneous Poisson point process with intensity~$\pi (t)$, then~$\hat U_n (\theta)$ are optimal estimating functions (i.e., most efficient) in the class of weighted estimating functions given by~\eqref{eq:approxscore} replacing~\eqref{eq:waage_weights} by any weight function~$w_i (t; \theta)$. This class includes the Horvitz-Thompson estimator under~$w(s; \theta) = 1$.
\end{prop}

\noindent Proposition~\ref{prop:optimal} ensures the only loss of statistical efficiency is due to subsampling and not using a suboptimal estimation procedure given subsampling.

Under assumption~\ref{assumption:truncation}, we next consider the statistical efficiency of our proposed estimator when compared to complete-data maximum likelihood estimation. While subsampling introduces additional variation, it may significantly reduce the overall computational burden. It is this trade-off that we next make precise. In particular, we consider the following choice of subsampling rate,~$\pi(t) = c \times h(t; \theta)$ for $c>0$. That is, the subsampling rate is proportional to the intensity function with time-independent constant~$c > 0$. Under this subsampling rate, the weight function~\eqref{eq:waage_weights} is equal to $c / (c+1)$. Under Lemma~\ref{lemma:simpleasym},
\[
\Xi (\theta) = \frac{c}{c+1} \int_0^\tau \frac{ h^{(1)} (t; \theta)
  h^{(1)} (t; \theta)^\top}{h (t; \theta)} dt = \frac{c}{c+1} \Sigma (\theta)
\]
where~$\Sigma(\theta)$ is the Fisher information of the complete-data maximum likelihood estimator.
Therefore the relative efficiency is~$c/(1+c)$. For an upper bound~$H = \max_{t \in (0,\tau)} h(t;\theta)$, if we set~$\pi (t) = c \times H$, then the relative efficiency can be lower bounded by~$c / (c+1)$.

@@

Recall sensor measurements occur on the order of multiple times per second.  Suppose the intensity rate is bounded above by $3$ and the unit time scale was hours. If we can subsample the data at~$30$ times per hour, then we have a lower bound on the efficiency of $0.909$. For a 4Hz sensor, this reduces the number of samples per hour from~$4 \times 60 \times 60 = 14,400$ per hour to on average $30$. While the computational complexity of logistic regression is linear in the number of samples, we get $480$ times reduction in the data size at the cost of a $0.909$ statistical efficiency. If we sample~$300$ times per hour, then the efficiency loss is only $0.999$, with a $48$ times reduction in data size. Table~\ref{tab:compvseff} provides additional examples for a 4Hz sensor.  The data reduction depends on the sensing rate; however, the lower bound on statistical efficiency does not.  This is because the subsampling rate only depends on the upper bound of the intensity function. That is, if the events are rare then the amount of data we need to subsample is greatly reduced with no impact to statistical efficiency.

@@

Recall theoretical results were proven under the assumption that there exists~$b^\star$ such that~$\beta(t) = \phi(t) {\bf b}^\star$. To make this assumption plausible, we set~$K_b$ large enough (but less than~$K_x$ to ensure identifiability) such that the spline basis expansion is sufficiently expressive. However, in practice, such a choice of~$K_b$ may lead to overfitting the data. We therefore, in the remainder of this paper, follow~\cite{GoldSmith2015} and induce smoothing by assuming that~$\{ b_k \}_{k=3}^{K_b} \sim N(0, \sigma_b^2 I)$. Incorporating penalization is simple due to the connection between with the score functions for logistic regression.

Due to the connection with generalized mixed effects models, we can leverage existing inferential machinery to obtain variance-covariance estimates of model parameters. That is, if~$\hat \Sigma_{bb}$ is the $K_b \times K_b$ deimnsional matrix obtained by plugging in the REML estimates of variance components into the formula for the variance of~$\hat b$, then the standard error for estimate at time~$t_0$ -- i.e., ~$\hat \beta (t_0) = \phi(t_0) \hat {\bf b}$ -- is given by~$\sqrt{ \phi (t_0 ) \hat \Sigma_{bb} \phi(t_0)^\top}$.  Then the approximate 95\% confidence interval can be constructed as~$\hat \beta (t_0) \pm 1.96 \sqrt{\phi (t_0) \hat \Sigma_{bb} \phi(t_0)^\top}$.
